The function--sheaf dictionary for smooth commutative group schemes over finite fields

### 2014-04-04 dictionary.tex
Added section called Group Cohomology. Started to thank Takashi.

%%%%%%%%%%%%%%%%% PACKAGES %%%%%%%%%%%%%%%%%
% Encodings
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
% AMS packages
\usepackage{amsmath, amsthm, amssymb, todonotes}
\usepackage[alphabetic]{amsrefs}
% Fonts
\usepackage{mathrsfs, yfonts}
\usepackage{geometry}
% TikZ
\usepackage{tikz}
\usetikzlibrary{shapes,arrows,calc,matrix}
\usepackage{tikz-cd}
% Hyperrefs
\usepackage{hyperref}

@@

%\begin{acknowledgements}
\subsection*{Acknowledgements}
We thank Masoud Kamgarpour, Pramod Achar, and Hadi Salmasian
for allowing us to hijack much of our Research in Teams meeting at BIRS into a discussion of
quasicharacter sheaves; their knowledge and help have been invaluable.
%We also thank Alessandra Bertrapelle and Cristian Gonz\'ales-Avil\'es for disabusing us of a misapprehension concerning the Greenberg realization functor and for pointing us to a proof of Lemma~\ref{lemma:wrK}.

@@

We will see that the Trace of Frobenius may not provide complete
information about isomorphism classes of quasicharacter sheaves on $G$ when $G$ is not connected.
Our main tool for understanding this phenomenon is the function
\begin{equation}\label{2cocycle}
\QCiso{G} \to H^0(\Weil{},H^2(G(\bFq),\EEx))
\end{equation}
defined below.
Let $\qcs{L} = (\gqcs{L},\mu,\phi)$ be a quasicharacter sheaf on $G$.
For each geometric point $x\in G(\bFq)$, choose a basis $\{ v_x \}$ for $\gqcs{L}_x$.
This choice determines a function $g : G(\bFq)\times G(\bFq) \to \EEx$ defined by
\[
\mu_{x,y}(v_{x+y}) = g(x,y) v_x \otimes v_y.
\]
Chasing through the diagram in Condition~\ref{QC.2} we see that
\[
g(x+y,z) g(x,y) = g(x,y+z) g(y,z),
\]
for all $x,y,z\in G(\bFq)$, so $g \in Z^2(G(\bFq),\EEx)$.
It is easy to see that the class of $g$ is independent of the choice made above;
let $g_\mathcal{L} \in H^2(G(\bFq),\EEx)$ the class of $g\in Z^2(G(\bFq),\EEx)$.
It is also easy to see that the class that $g_\mathcal{L} = g_\mathcal{L'}$ when $\qcs{L} \iso \qcs{L'}$. This defines
\[
\QCiso{G} \to H^2(G(\bFq),\EEx).
\]
We claim that $g_\qcs{L}$ is also fixed by the action of $\Weil{}$ on $H^2(G(\bFq),\EEx)$
inherited from the action of $\Weil{}$ on $G(\bFq)$.
To see this, we return to the choice $\{ v_x \in \gqcs{L}_x^\times \tq x\in G(\bFq) \}$
and define $f : G(\bFq) \to \EEx$ by
\[
\phi_x(v_{\Frob{G}(x)}) = f(x) v_x.
\]
Chasing through the diagram in Condition~\ref{QC.3} we see that
\begin{equation}\label{nohom}
f(x+y) = f(x) f(y) \frac{(\,^{\Frob{}}g)(x,y)}{g(x,y)}
\end{equation}
for all $x,y \in G(\bFq)$. Thus, the class $\,^{\Frob{}}g_\qcs{L}$ of $\,^{\Frob{}}g$ coincides
with the class $g_\qcs{L}$ of $g$. This concludes the definition of
\[
\QCiso{G} \to H^0(\Weil{},H^2(G(\bFq),\EEx)).
\]

@@

If $G$ is a connected and of finite type, 
then $H^2(G(\bFq),\EEx)$ is trivial \todo{Reference?} so all quasicharacter sheaves on connected commutative algebraic groups $G$ are visible; 
 by contrast, if $G$ is etale then $H^2(G(\bFq),\EEx)$ may be non-trivial and $G$ may admit invisible quasicharacters sheaves.
We will return to the latter case in Section~\ref{sec:etale}.
In the meantime, we end this section with a simple property of \eqref{2cocycle}.

\begin{lemma}
The function defined in \eqref{2cocycle} is a homomorphism:
\[
g_{\qcs{L}\otimes \qcs{L'}} = g_{\qcs{L}} \, g_\qcs{L'}
\]
for every $\qcs{L}, \qcs{L'} \in \QC(G)$.
\end{lemma}

\todo{Proof missing.}

@@

\begin{lemma}\label{lem:etale-2cocycle}
    If $G$ is an \'etale commutative group scheme over $\Fq$
    then the group homomorphism
    \[
    \QCiso{G} \to \Hh^0(\Weil{}, \Hh^2(G(\bFq),\EEx)),
    \]
    defined in Section~\ref{sec:invisible}, 
    is surjective with a canonical section.
\end{lemma}

\begin{proof}
Pick $[g]\in \Hh^0(\Weil{}, \Hh^2(G(\bFq),\EEx))$. Then $g\in Z^2(G(\bFq),\EEx)$ is fixed by Frobenius, so $\,^{\Frob{}}g = g h$ for some $h\in B^2(G(\bFq),\EEx)$. Accordingly, $h(x,y) = \frac{f(x+y)}{f(x) f(y)}$ for some $f : G(\bFq) \to \EEx$, so
\begin{equation}\label{gf}
g(\Frob{}(x),\Frob{}(y)) = g(x,y) \frac{f(x+y)}{f(x) f(y)}
\end{equation}
for all $x,y\in G(\bFq)$.
Set $\gqcs{L}_x = \EE$ for every $x\in G(\bFq)$ and define $\mu_{x,y} : \gqcs{L}_{x+y} \to \gqcs{L}_x \otimes \gqcs{L}_y$ by $\mu_{x,y}(1) = g(x,y) 1 \otimes 1$.
Since $g\in Z^2(G(\bFq),\EEx)$, the indexed set of isomorphisms $\{ \mu_{x,y} \tq x,y \in G(\bFq) \}$ satisfies \ref{qc.2}.
Define $\phi_x : \gqcs{L}_{\Frob{}(x)}\to \gqcs{L}_x$ by $\phi_x(1) = f(x)$.
Then set of isomorphisms $\{ \phi_{x} \tq x \in G(\bFq) \}$ satisfies \ref{qc.3} by \eqref{gf}. In this way, the cocycle $g\in Z^0(\Weil{}, Z^2(G(\bFq),\EEx))$ defines a quasicharacter sheaf $\qcs{L}$. It is clear that $\qcs{L}$ maps to $[g] \in \Hh^0(\Weil{}, \Hh^2(G(\bFq),\EEx))$ under \eqref{2cocycle}.\todo{Canonical section?} 
\end{proof}

@@

\begin{lemma}\label{lem:etale-trace}
If $G$ is an \'etale commutative group scheme over $\Fq$
then for every the $\chi \in G(\Fq)^*$ there is a visible quasicharacter sheaf $\qcs{L}$ on $G$ such that $\trFrob{\qcs{L}} = \chi$.
In particular, Trace of Frobenius
\[
\TrFrob{G} : \QCiso{G} \to G(\Fq)^*,
\]
is surjective, even when restricted to visible quasicharacter sheaves.
\end{lemma}

\begin{proof}
Pick $\chi \in G(\Fq)^*$.
By Lemma~\ref{lem:dual-inv}, there is some $f\in G(\bFq)^*$ such that $f\vert_{G(\Fq)^*} = \chi$.
Set $\gqcs{L}_x = \EE$ for every $x\in G(\bFq)$.
Define $\mu_{x,y} : \gqcs{L}_{x+y} \to \gqcs{L}_x\otimes \gqcs{L}_y$ by $\mu_{x,y}(1) = 1 \otimes 1$.
Define $\phi_{x} : \gqcs{L}_{\Frob{}(x)} \to \gqcs{L}_x$ by $\phi_{x}(1) = f(x)$.
Since $f : G(\bFq) \to \EEx$ is a group homomorphism,
condition \eqref{nohom} is satisfied with $g =1$, the trivial cocycle in $Z^2(G(\bFq),\EEx)$.
It is now clear that we have defined a \emph{visible} quasicharacter sheaf $\qcs{L}$ with $\trFrob{\qcs{L}} = \chi$.\todo{Can I improve this proof to show that the kernel of Trace of Frobenius is exactly $\Hh^0(\Weil{},\Hh^2(G(\bFq),\EEx))$? Perhaps . . . }
\end{proof}

@@

\begin{proof}
The Hochschild-Serre spectral sequence (for finitely generated abelian groups!) provides decomposition 
\[
\Hh^2(G(\bFq),\EEx)  \iso \Hh^1(\Weil{}, \Hh^1(G(\bFq)\rtimes \Weil{},\EEx)) \oplus \Hh^0(\Weil{}, \Hh^2(G(\bFq),\EEx))
\]
and Lemmas~\ref{lem:etale-2cocycle} and \ref{lem:etale-trace} provide the maps from $\QCiso{G}$ pictured below:
\[
\begin{tikzcd}
\ &&  \QCiso{G} \arrow[swap]{dl}{\TrFrob{G}} \arrow{dr}{ S_{G} } && \\
0 \arrow{r}{} &  \Hh^1(\Weil{}, \Hh^1(G(\bFq),\EEx)) \arrow{r}{} & \Hh^2(G(\bFq)\rtimes\Weil{},\EEx) \arrow{r}{} & \Hh^0(\Weil{}, \Hh^2(G(\bFq),\EEx)) \arrow{r}{} & 0.
\end{tikzcd}
\]

\todo{Clifton, you are here.}
\end{proof}

>>>>

%%%%%%%%%%%%%%%%% PACKAGES %%%%%%%%%%%%%%%%%
% Encodings
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
% AMS packages
\usepackage{amsmath, amsthm, amssymb, todonotes}%showkeys}
\usepackage[alphabetic]{amsrefs}
% Fonts
\usepackage{mathrsfs, yfonts}
\usepackage{geometry}
% TikZ
\usepackage{tikz}
\usetikzlibrary{shapes,arrows,calc,matrix}
\usepackage{tikz-cd}
% Hyperrefs
\usepackage{hyperref}

@@

%\begin{acknowledgements}
\subsection*{Acknowledgements}
We thank Masoud Kamgarpour, Pramod Achar, and Hadi Salmasian
for allowing us to hijack much of our Research in Teams meeting at BIRS into a discussion of
quasicharacter sheaves; their knowledge and help have been invaluable.
We also thank Takashi Suzuki, who

@@

We will see that the Trace of Frobenius may not provide complete
information about isomorphism classes of quasicharacter sheaves on $G$ when $G$ is not connected.
Our main tool for understanding this phenomenon is the function
\begin{equation}\label{2cocycle}
S_{G}: \QCiso{G} \to H^0(\Weil{},H^2(G(\bFq),\EEx))
\end{equation}
defined below.
Let $\qcs{L} = (\gqcs{L},\mu,\phi)$ be a quasicharacter sheaf on $G$.
For each geometric point $x\in G(\bFq)$, choose a basis $\{ v_x \}$ for $\gqcs{L}_x$.
This choice determines a function $g : G(\bFq)\times G(\bFq) \to \EEx$ defined by
\[
\mu_{x,y}(v_{x+y}) = g(x,y) v_x \otimes v_y.
\]
Chasing through the diagram in Condition~\ref{QC.2} we see that
\[
g(x+y,z) g(x,y) = g(x,y+z) g(y,z),
\]
for all $x,y,z\in G(\bFq)$, so $g \in Z^2(G(\bFq),\EEx)$.
It is easy to see that the class of $g$ is independent of the choice made above;
let $g_\mathcal{L} \in H^2(G(\bFq),\EEx)$ the class of $g\in Z^2(G(\bFq),\EEx)$.
It is also easy to see that the class that $g_\mathcal{L} = g_\mathcal{L'}$ when $\qcs{L} \iso \qcs{L'}$. This defines
\[
\QCiso{G} \to H^2(G(\bFq),\EEx).
\]
We claim that $g_\qcs{L}$ is also fixed by the action of $\Weil{}$ on $H^2(G(\bFq),\EEx)$
inherited from the action of $\Weil{}$ on $G(\bFq)$.
To see this, we return to the choice $\{ v_x \in \gqcs{L}_x^\times \tq x\in G(\bFq) \}$
and define $f : G(\bFq) \to \EEx$ by
\[
\phi_x(v_{\Frob{G}(x)}) = f(x) v_x.
\]
Chasing through the diagram in Condition~\ref{QC.3} we see that
\begin{equation}\label{nohom}
f(x+y) = f(x) f(y) \frac{g(\Frob{G}(x),\Frob{G}(y))}{g(x,y)}
\end{equation}
for all $x,y \in G(\bFq)$. Thus, the class $\,^{\Frob{}}g_\qcs{L}$ of $\,^{\Frob{}}g$ coincides
with the class $g_\qcs{L}$ of $g$. This concludes the definition of
$S_{G}: \QCiso{G} \to H^0(\Weil{},H^2(G(\bFq),\EEx))$.

@@

If $G$ is a connected and of finite type, 
then $H^2(G(\bFq),\EEx)$ is trivial \todo{Reference?} so all quasicharacter sheaves on connected commutative algebraic groups $G$ are visible; 
 by contrast, if $G$ is etale then $H^2(G(\bFq),\EEx)$ may be non-trivial and $G$ may admit invisible quasicharacters sheaves.
We will return to the latter case in Section~\ref{sec:etale}.
In the meantime, we end this section with a simple property of $S_{G}: \QCiso{G} \to H^0(\Weil{},H^2(G(\bFq),\EEx))$.

\begin{lemma}
The function \[S_{G}: \QCiso{G} \to H^0(\Weil{},H^2(G(\bFq),\EEx))\] is a group homomorphism.
\end{lemma}

\begin{proof}
We must show:
\[
g_{\qcs{L}\otimes \qcs{L'}} = g_{\qcs{L}} \, g_\qcs{L'}
\]
for every $\qcs{L}, \qcs{L'} \in \QC(G)$.\todo{Under construction.}
\end{proof}

@@

\section{Group Cohomology}

\begin{lemma}\label{lem:etale-2cocycle}
    If $G$ is an \'etale commutative group scheme over $\Fq$
    then the group homomorphism
    \[
    S_{G} : \QCiso{G} \to \Hh^0(\Weil{}, \Hh^2(G(\bFq),\EEx)),
    \]
    defined in Section~\ref{sec:invisible}, is surjective.\todo{Kernel? Canonical section?}
\end{lemma}

\begin{proof}
Pick $[g]\in \Hh^0(\Weil{}, \Hh^2(G(\bFq),\EEx))$. Then $g\in Z^2(G(\bFq),\EEx)$ is fixed by Frobenius, so $\,^{\Frob{}}g = g h$ for some $h\in B^2(G(\bFq),\EEx)$. Accordingly, $h(x,y) = \frac{f(x+y)}{f(x) f(y)}$ for some $f : G(\bFq) \to \EEx$, so
\begin{equation}\label{gf}
g(\Frob{}(x),\Frob{}(y)) = g(x,y) \frac{f(x+y)}{f(x) f(y)}
\end{equation}
for all $x,y\in G(\bFq)$.
Set $\gqcs{L}_x = \EE$ for every $x\in G(\bFq)$ and define $\mu_{x,y} : \gqcs{L}_{x+y} \to \gqcs{L}_x \otimes \gqcs{L}_y$ by $\mu_{x,y}(1) = g(x,y) 1 \otimes 1$.
Since $g\in Z^2(G(\bFq),\EEx)$, the indexed set of isomorphisms $\{ \mu_{x,y} \tq x,y \in G(\bFq) \}$ satisfies \ref{qc.2}.
Define $\phi_x : \gqcs{L}_{\Frob{}(x)}\to \gqcs{L}_x$ by $\phi_x(1) = f(x)$.
Then set of isomorphisms $\{ \phi_{x} \tq x \in G(\bFq) \}$ satisfies \ref{qc.3} by \eqref{gf}. In this way, the cocycle $g\in Z^0(\Weil{}, Z^2(G(\bFq),\EEx))$ defines a quasicharacter sheaf $\qcs{L}$. It is clear that $\qcs{L}$ maps to $[g] \in \Hh^0(\Weil{}, \Hh^2(G(\bFq),\EEx))$ under \eqref{2cocycle}.
\end{proof}

@@

\begin{lemma}\label{lem:etale-trace}
If $G$ is an \'etale commutative group scheme over $\Fq$
then for every the $\chi \in G(\Fq)^*$ there is a visible quasicharacter sheaf $\qcs{L}$ on $G$ such that $\trFrob{\qcs{L}} = \chi$.
In particular, Trace of Frobenius
\[
\TrFrob{G} : \QCiso{G} \to G(\Fq)^*,
\]
is surjective, even when restricted to visible quasicharacter sheaves.\todo{Kernel? Canonical section?}
\end{lemma}

\begin{proof}
Pick $\chi \in G(\Fq)^*$.
By Lemma~\ref{lem:dual-inv}, there is some $f\in G(\bFq)^*$ such that $f\vert_{G(\Fq)^*} = \chi$.
Set $\gqcs{L}_x = \EE$ for every $x\in G(\bFq)$.
Define $\mu_{x,y} : \gqcs{L}_{x+y} \to \gqcs{L}_x\otimes \gqcs{L}_y$ by $\mu_{x,y}(1) = 1 \otimes 1$.
Define $\phi_{x} : \gqcs{L}_{\Frob{}(x)} \to \gqcs{L}_x$ by $\phi_{x}(1) = f(x)$.
Since $f : G(\bFq) \to \EEx$ is a group homomorphism,
condition \eqref{nohom} is satisfied with $g =1$, the trivial cocycle in $Z^2(G(\bFq),\EEx)$.
It is now clear that we have defined a \emph{visible} quasicharacter sheaf $\qcs{L}$ with $\trFrob{\qcs{L}} = \chi$.
\end{proof}

@@

\begin{proof}
By Lemma~\ref{lem:dual-inv}, $G(\Fq)^* \iso \Hh^1(\Weil{},\Hh^1(G(\bFq),\EEx))$. Composing this isomorphism with the Trace of Frobenius gives 
\[
T_{G}: \QCiso{G}\to  \Hh^1(\Weil{},\Hh^1(G(\bFq),\EEx)),
\]
which is surjective, Lemma~\ref{lem:etale-trace}. 
Combining this with 
\[
S_{G} : \QCiso{G} \to \Hh^0(\Weil{}, \Hh^2(G(\bFq),\EEx))
\]
gives
\[
\QCiso{G} \mathop{\longrightarrow}\limits^{T_{G}\oplus S_{G}} \Hh^1(\Weil{},\Hh^1(G(\bFq),\EEx))\oplus \Hh^0(\Weil{}, \Hh^2(G(\bFq),\EEx)).
\]
The Hochschild-Serre spectral sequence (for finitely generated abelian groups!) provides  
\[
\Hh^2(G(\bFq)\rtimes \Weil{},\EEx)  \iso \Hh^1(\Weil{}, \Hh^1(G(\bFq),\EEx)) \oplus \Hh^0(\Weil{}, \Hh^2(G(\bFq),\EEx)).
\]
Composing these last two maps defines
\begin{equation}
\QCiso{G}\to \Hh^2(G(\bFq)\rtimes \Weil{},\EEx).
\end{equation}
To see that this homomorphism is an isomorphism,\todo{Clifton, you are here.}
\end{proof}
